MLPPI Wizard: An Automated Multi-level Partitioning Tool on Analytical Workloads

### 2017-03-14 mlppi-comsis17/paper.tex
Done before the Optimized Phase

\institute{Department of Computer Science, The University of Arizona, Tucson, AZ 85721, USA\\ 
  \email{yksuh@cs.arizona.edu}
  \and
  Teradata Corporation, El Segundo, CA 90245, USA\\
  \email{\{alain.crolotte,pekka.kostamaa\}@teradata.com}
%  \and
%  Korea Institute of Science and Technology Information, Daejeon 34141 Republic of Korea\\ 
%  \email{yksuh@kisti.re.kr}
}

@@

By and large, overall I/O cost may be increased by the merge, 
in that the full scan on a partition is performed to retrieve all the rows potentially matching a given query. 
In the case of a merged partition, we may read the non-qualifying rows 
that would not be seen before the merge, thereby paying more I/O to answer 
the query. To minimize the merge overhead, we pick up the range pair incurring 
the least I/O increase in a heuristic fashion. 
To choose the desirable range pair for a merge, 
the tool leverages the {\em scan cost} of a query on a range pair.

@@

\begin{algorithm}[t]
\caption{Scan Cost Computation}
\label{algo:comp_scan_cost}
\SetKwInOut{Input}{input}
\SetKwInOut{Output}{output}
\SetKw{func}{Function}
\SetKwBlock{begin}{}{end}
{
\Input{$q$ (query), $rp$ (range pair), $M$ (query-to-range-set map)}
\Output{Scan cost for answering $q$ when considering a merge of $rp$}
	$r_{m}$ $\leftarrow$ Consolidate $rp$ \;
	$L$ $\leftarrow$ Copy the range set mapped to $q$ from $M$\; 
	Delete ranges in $rp$ from and insert $r_{m}$ into $L$. \\
	$s$ $\leftarrow$ ``{\tt SELECT * FROM FACT$\_$TABLE WHERE }'' \;
	\ForEach{range $r$ $\in$ $L$}{	
		Restore a predicate $p$ from $r$. \\
		Add $p$ to {\tt WHERE} clause of $s$.
	}
	Make an API call with $s$ to the server. \\
	Extract the spool size (in bytes) from the result. \\
	$blc$ $\leftarrow$ Calculate the block counts from the spool size. \\
	Update $q$'s scan cost to $blc$. \\
	{\bf return} $blc$ \;
}
\end{algorithm}

If $q$ turns out to be affected by the merge of a range pair, 
then the range set associated with $q$ in $M$ is temporarily updated 
by removing the parent ranges and adding the merged range. 
The altered range set is remapped to $q$ in $M$ and then 
translated to the equivalent predicates. Then {\it CP} for $s$ can be built based on the predicates.
Unless the merge range pair influences $q$, 
we can simply derive {\it CP} from the existing ranges mapped to $q$.

To help understand the scan cost construction, let us 
consider a range pair ($rp$) of \hbox{$R$[2, 2]} and $R$[2, 3] in the running example. 
$rp$ produces the merged range, \range{25, 35}, \hbox{under} {\tt LO\_QUANTITY}. 
The merge by $rp$ affects both $q$1 and $q$2 in the workload.

Thus, the range sets of $q$1 and $q$2 on {\tt LO\_QUANTITY} are 
altered to \{\openrange{\infty, 25}, \range{25, 35}\} and \{\range{25, 35}\}, 
respectively. Of course, no change is made to the existing range sets 
of the queries on {\tt LO\_DISCOUNT}. 
Therefore, the corresponding scan cost queries for $q$1 and $q$2 
can be built as below:
\begin{center}
\begin{tabular}{rl}
$\Sone$:	& {\tt SELECT * FROM LINEORDER l} \\
		& {\tt WHERE ((l.LO\_DISCOUNT = 1)} \\
		& {\tt OR (l.LO\_DISCOUNT >= 4} \\ 
		& {\tt AND l.LO\_DISCOUNT <= 5))} \\
		& {\tt AND l.LO\_QUANTITY <= 35}  \\ 
$\Stwo$:	& {\tt SELECT * FROM LINEORDER l} \\
        & {\tt WHERE l.LO\_DISCOUNT >= 7} \\
        & {\tt AND (l.LO\_QUANTITY >= 25 AND} \\ 
        & {\tt l.LO\_QUANTITY <= 35)}.
\end{tabular}
\end{center}

It is possible to combine all the (consecutive) ranges in the (temporarily) altered set of $q$1, 
and hence, a single predicate ({\tt l.LO\_QUANTITY <= 35}) can be built.

The wizard sends such a scan cost query to the server via an API call, 
and it extracts from the server's response the {\em spool size} (in bytes), or the size of scan query result. 
Then, the tool computes as scan cost the {\em block counts} using the spool size. 
If the merge of a range pair does not impact on any query in the workload, 
the existing (previously computed) scan cost of queries 
will be reused for the range pair. 
That is, the wizard will only recompute 
the scan cost of a query {\em iff} the query is affected by 
the merge of two consecutive ranges.

In this way, the weighted sum of scan cost of queries, $T_{s}$ can be computed 
for each range pair. 
$T_{s}$ can be defined as follows:
\begin{center}
\begin{tabular}{c} 
$T_{s}$ = $\sum_{i=1}^{n}$($sc_{i}${$\cdot$}{$w_{i}$}),
\end{tabular}
\end{center}
where $n$ is the number of queries in $Q$, $sc_{i}$ is the scan cost of a query $q_{i}$ in $Q$, and 
$w_{i}$ denotes the (\hbox{non-negative}) weight associated with $q_{i}$.

@@

Note that we see that $q$2 may have the most customized partition based on the updated $M$. 
Only the single partition formed by ($R$[1, 3]$\cap${$R$[2, 2]}) is sufficient 
to retrieve all the qualifying 
rows for $q$2; thus, the other partitions can be simply eliminated. 
In the meantime, to answer $q$1, we need to read the four partitions formed by 
the top two ranges of each field. 
Unfortunately, because the partitions formed by (({$R$[1, 1]{$\cup$}$R$[1, 2]})$\cap$($R$[2, 2]) 
contain the non-qualifying rows for $q$1, $R$ cannot provide $q$1 with as much benefit as $q$2. 
However, $R$ can be a good compromise to satisfy both queries, in that 
the MLPPI derived by $R$ can potentially minimize the total execution cost 
of $Q$.

@@

Algorithm~\ref{algo:init_phase} represents the initial phase 
that has been proposed so far. 
The algorithm starts with a query-to-range set map ($M$) and an input range set ($R$). 
Until the number of the partitions in $R$ falls below the partition limit threshold, 
the algorithm determines the range pair yielding the least total execution cost ($T_{s}$) 
and updates $M$ and $R$. At the end, the algorithm produces a final recommendation based on $R$.
%\vspace{-.1in}

Now that the total partition counts ($4${$\times$}$3$=$12$) by $R$ falls  
below the assumed limit ($15$), the wizard 
proceeds to the optimized phase with $R$ without repeating the initial phase.

@@

%We verified the effectiveness of the tool by running it on a workload based
In this regard, using $R$ we have an initial MLPPI recommendation for the {\tt LINEORDER} fact table.
The recommendation can be used sufficiently, but 
having fewer partitions, induced by further merges, 
can enhance the overall workload performance for the following reasons. 
First, multiple file contexts by many partitions can incur 
huge overhead that impacts on the query optimizer. 
There also exists an operational threshold 
that the optimizer can handle the maximum number of the partitions at a time.
Therefore, further reducing partitions greatly helps the optimizer to manage these partitions.

@@

In the running example, suppose that in the first round, ranges $R$[1, 1] and $R$[1, 2] are chosen for a merge. 
Then, the wizard produces a merged range \range{1, 5} \hbox{under} {\tt LO\_DISCOUNT} and proceeds to the next round. 
If a range pair selected for the subsequent merge fails to improve $T_{p}$, 
then, the tool exits the optimized phase and \hbox{produces} the final MLPPI recommendation for the given workload $Q$ 
as shown in Figure~\ref{fig:mlppi}.

\begin{figure}[h]
\begin{center}
\begin{tabular}{|l|} \hline
{\tt ALTER TABLE LINEORDER} \\
{\tt MODIFY PRIMARY INDEX} \\
{\tt PARTITION BY(} \\
\hspace{.1in}{\tt CASE\_N(} \\
	\hspace{.2in}{\tt LO\_DISCOUNT $\geq$ 1 AND LO\_DISCOUNT $\leq$ 5,} \\
	\hspace{.2in}{\tt LO\_DISCOUNT $\geq$ 7,} \\ 
	\hspace{.2in}{\tt NO CASE OR UNKNOWN),} \\
\hspace{.1in}{\tt CASE\_N(} \\
	\hspace{.2in}{\tt LO\_QUANTITY $<$ 25,} \\
	\hspace{.2in}{\tt LO\_QUANTITY $\geq$ 25 AND LO\_QUANTITY $\leq$ 35,} \\
	\hspace{.2in}{\tt NO CASE OR UNKNOWN)} \\ 
{\tt );} \\ \hline
\end{tabular}
\end{center}
\caption{An MLPPI Recommendation for Workload $Q$\label{fig:mlppi}}
\end{figure}

@@

{\bf Comment:} In practice, our experiments showed that the bound 
$\BigO$($N${$\cdot$}$C${$\cdot$}$V$) is overly pessimistic.

@@

\begin{lemma}
\label{lemma:init_phase}  
The running time complexity for the initial or optimized phases is 
$\BigO$($M^{2}${$\cdot$}$N$), 
where $M$ is the number of range pairs, and $N$ is the number of queries in 
a given workload. 
\end{lemma}

@@

Our experiments showed that $\BigO$($M^{2}${$\cdot$}$N$), the running time complexity, 
is overly pessimistic. 
In \hbox{practice}, the number of calls made was much fewer than the \hbox{theoretic} bound, 
since it was very rare for a range pair to be associated with all queries in a workload. 
In \hbox{Section}~\ref{sec:stat} we will show how many API calls take place on the workloads 
used for evaluation.

@@

The total number of iterations in the initial phases 
tended to be proportional to the input \hbox{partitions} 
\hbox{generated} from the workloads. 
Note that there \hbox{happened} only one iteration in the optimized phase. 
The reason was that since the partitions produced by the \hbox{initial} phase were customized enough, 
no more \hbox{iterations} were \hbox{necessary} for further optimization.

@@

\section*{Acknowledgments} 
This work was carried out while the first author visited Teradata. 
The first author is now at Korea Institute of Science and Technology Information (yksuh@kisti.re.kr).

>>>>

\institute{Department of Computer Science, The University of Arizona, AZ 85721, USA\footnote{Currently, he is at Korea Institute of Science and Technology Information (KISTI). This work was conducted while he interned at Teradata during his PhD in the University of Arizona.}\\ 
  %\email{yksuh@cs.arizona.edu}
  \email{yksuh@cs.arizona.edu}
  \and
  Teradata Corporation, El Segundo, CA 90245, USA\\
  \email{\{alain.crolotte,pekka.kostamaa\}@teradata.com}
}

@@

Overall, I/O cost may be increased by the merge, 
in that the full scan on a partition is performed to retrieve all the rows potentially matching a given query. 
In the case of a merged partition, we may read the non-qualifying rows 
that would not be seen before the merge, thereby paying more I/O to answer 
the query. To minimize the merge overhead, we pick up the range pair incurring 
the least I/O increase in a heuristic fashion. 
To choose a range pair for a merge, 
we calculate the {\em scan cost} of a query on the merged range pair.

@@

\begin{algorithm}[t]
\caption{Scan Cost Computation}
\label{algo:comp_scan_cost}
\SetKwInOut{Input}{input}
\SetKwInOut{Output}{output}
\SetKw{func}{Function}
\SetKwBlock{begin}{}{end}
{
\Input{$q$ (query), $rp$ (range pair), $M$ (query-to-range-set map)}
\Output{Scan cost for answering $q$ when considering a merge of $rp$}
	$r_{m}$ $\leftarrow$ Consolidate $rp$ \;
	$L$ $\leftarrow$ Copy the range set mapped to $q$ from $M$\; 
	Delete ranges in $rp$ from and insert $r_{m}$ into $L$. \\
	$s$ $\leftarrow$ ``{\tt SELECT * FROM FACT$\_$TABLE WHERE }'' \;
	\ForEach{range $r$ $\in$ $L$}{	
		Reconstruct a predicate $p$ from $r$. \\
		Add $p$ to {\tt WHERE} clause of $s$.
	}
	Make an API call with $s$ to the server. \\
	Extract the spool size (in bytes) from the result. \\
	$blc$ $\leftarrow$ Calculate the block counts from the spool size. \\
	Update $q$'s scan cost to $blc$. \\
	{\bf return} $blc$ \;
}
\end{algorithm}

If $q$ turns out being affected by the merge of a range pair, 
then the range set associated with $q$ in $M$ is temporarily updated 
by removing the parent ranges and adding the merged range. 
The altered range set is remapped to $q$ in $M$ and then 
translated to the equivalent predicates. Then {\it CP} for $s$ can be built based on the predicates.
Unless the merge range pair influences $q$, 
we can simply derive {\it CP} from the existing ranges mapped to $q$.

To see in the Algorithm~\ref{algo:comp_scan_cost} how a scan cost query is built, 
consider a range pair ($rp$) of \hbox{$R$[2, 2]} and $R$[2, 3] in the running example. 
$rp$ produces the merged range, \range{25, 35}, \hbox{under} {\tt LO\_QUANTITY}. 
The merge by $rp$ affects both $q$1 and $q$2 in the workload. 
Thus, the range sets of $q$1 and $q$2 on {\tt LO\_QUANTITY} are 
correspondingly altered to \{\openrange{\infty, 25}, \range{25, 35}\} and \{\range{25, 35}\}, 
respectively. Of course, no change is made to the existing range sets 
of the queries on {\tt LO\_DISCOUNT}. 
Therefore, the corresponding scan cost queries for $q$1 and $q$2 
can be built as shown below:
\vspace{-.08in}
\begin{center}
\begin{tabular}{rl}
$\Sone$:	& {\tt SELECT * FROM LINEORDER l} \\
		& {\tt WHERE ((l.LO\_DISCOUNT = 1)} \\
		& {\tt OR (l.LO\_DISCOUNT >= 4} \\ 
		& {\tt AND l.LO\_DISCOUNT <= 5))} \\
		& {\tt AND l.LO\_QUANTITY <= 35}  \\ 
$\Stwo$:	& {\tt SELECT * FROM LINEORDER l} \\
        & {\tt WHERE l.LO\_DISCOUNT >= 7} \\
        & {\tt AND (l.LO\_QUANTITY >= 25 AND} \\ 
        & {\tt l.LO\_QUANTITY <= 35)}.
\end{tabular}
\end{center} 
\vspace{-.08in}
\noindent Since the altered ranges (\openrange{\infty, 25} and \range{25, 35}) of $q$1 
are consecutive, we can simply build the merged, single predicate ({\tt l.LO\_QUANTITY <= 35}) for $\Sone$.

In turn, the wizard sends each scan cost query to the server through an API call. 
The tool extracts from the server's response the {\em spool size} (in bytes), which is 
the result size of the sent scan cost query, and then 
calculates as scan cost the {\em block counts} ($blc$) using the spool size. 
If the merge of a range pair does not impact on any query in the workload, 
the existing (previously computed) scan cost of queries is reused for the range pair. 
In other words, the wizard only recomputes the scan cost of a query {\em only if} 
the query is affected by the merge of two consecutive ranges. 
The weighted sum of scan cost of queries, $T_{s}$ can be computed 
for each range pair. 
$T_{s}$ can be defined as follows:
\begin{center}
\begin{tabular}{c} 
$T_{s}$ = $\sum_{i=1}^{n}$($sc_{i}${$\cdot$}{$w_{i}$}),
\end{tabular}
\end{center}
where $n$ is the number of queries in $Q$, $sc_{i}$ is the scan cost of a query $q_{i}$ in $Q$, and 
$w_{i}$ denotes the (\hbox{non-negative}) weight associated with $q_{i}$.

@@

Note that we see that $q$2 may have the most customized partition based on the updated $M$. 
Only the single partition formed by ($R$[1, 3]$\cap${$R$[2, 2]}) is sufficient 
to retrieve all the qualifying 
rows for $q$2; thus, the other partitions can be simply eliminated. 
In the meantime, to answer $q$1, we need to read the four partitions formed by 
the top two ranges of each field. 
As the partitions formed by (({$R$[1, 1]{$\cup$}$R$[1, 2]})$\cap$($R$[2, 2]) 
contain the non-qualifying rows for $q$1, unfortunately, 
$R$ cannot provide $q$1 with as much benefit as $q$2. 
$R$, however, can be a good compromise to satisfy both queries, in that 
the MLPPI derived by $R$ can potentially minimize the total execution cost of $Q$.

@@

Algorithm~\ref{algo:init_phase} represents the initial phase 
that has been described so far. 
Given a \linebreak \hbox{query-to-range} set map ($M$) and an input range set ($R$), 
the algorithm determines which range pair to yield the least total execution cost ($T_{s}$) 
and updates $M$ and $R$ using the chosen range pair, 
until the number of the partitions by $R$ falls below a pre-defined partition limit count.
Finally, the algorithm produces the final recommendation based on $R$.

In the running example the total partition counts ($4${$\times$}$3$=$12$) by $R$ is under 
the assumed limit ($15$), which meets the partitioning limit. 
Hence, the wizard can enter the optimized phase with $R$, 
without repeating the initial phase.

@@

%We verified the effectiveness of the tool by running it on a workload based
After passing the initial phase, 
we now have an initial MLPPI recommendation based on $R$, for the {\tt LINEORDER} fact table.
The initial recommendation can be used sufficiently.

But via additional merges we may further make fewer partitions, 
which actually can enhance the overall workload performance for the following reasons. 
First, multiple file contexts by many partitions can incur 
huge overhead that impacts on the query optimizer. 
There also exists an operational threshold 
that the optimizer can handle the maximum number of the partitions at a time.
Therefore, further reducing partitions greatly helps the optimizer to manage these partitions.

@@

In the running example, suppose that in the first round, ranges $R$[1, 1] and $R$[1, 2] are chosen for a merge. 
Then, the wizard produces a merged range \range{1, 5} \hbox{under} {\tt LO\_DISCOUNT} and proceeds to the next round. 
If a range pair selected for the subsequent merge fails to improve $T_{p}$, 
then, the tool exits the optimized phase and \hbox{produces} the final MLPPI recommendation for the given workload $Q$ 
as shown in Figure~\ref{fig:mlppi}. 
\vspace{-.25in}
\begin{figure}[h]
\begin{center}
\begin{tabular}{|l|} \hline
{\tt ALTER TABLE LINEORDER} \\
{\tt MODIFY PRIMARY INDEX} \\
{\tt PARTITION BY(} \\
\hspace{.1in}{\tt CASE\_N(} \\
	\hspace{.2in}{\tt LO\_DISCOUNT $\geq$ 1 AND LO\_DISCOUNT $\leq$ 5,} \\
	\hspace{.2in}{\tt LO\_DISCOUNT $\geq$ 7,} \\ 
	\hspace{.2in}{\tt NO CASE OR UNKNOWN),} \\
\hspace{.1in}{\tt CASE\_N(} \\
	\hspace{.2in}{\tt LO\_QUANTITY $<$ 25,} \\
	\hspace{.2in}{\tt LO\_QUANTITY $\geq$ 25 AND LO\_QUANTITY $\leq$ 35,} \\
	\hspace{.2in}{\tt NO CASE OR UNKNOWN)} \\ 
{\tt );} \\ \hline
\end{tabular}
\end{center}
\vspace{-.2in}
\caption{The Final MLPPI Recommendation for the Given Workload $Q$\label{fig:mlppi}}
\vspace{-.2in}
\end{figure}

@@

But we empirically observed that the bound of $\BigO$($N${$\cdot$}$C${$\cdot$}$V$) 
was overly pessimistic.

@@

\begin{lemma}
\label{lemma:init_phase}  
The running time complexity for the initial or optimized phases is 
$\BigO$($M^{2}${$\cdot$}$N$), 
where $M$ is the number of range pairs, and $N$ is the number of queries in a workload. 
\end{lemma}

@@

Again, our experiments showed that $\BigO$($M^{2}${$\cdot$}$N$), the running time complexity, 
was overly pessimistic. Specifically, 
the number of calls made in our experiments was much fewer than that bound, as 
it was almost rare for a range pair to involve all queries in a workload 
(as seen in Table~\ref{tab:workload_stat}).

In the following, we evaluate the proposed algorithm.

@@

The total number of iterations in the initial phases 
tended to be proportional to the input \hbox{partitions} 
\hbox{generated} from the workloads. 
Note that there \hbox{happened} only one iteration in the optimized phase. 
The reason was that as the partitions produced by the \hbox{initial} phase were customized enough, 
no more \hbox{iterations} were \hbox{necessary} for further optimization.

@@

%\section*{Acknowledgments} 
%This work was carried out while the first author visited Teradata. 
%The first author is now at Korea Institute of Science and Technology Information (yksuh@kisti.re.kr).
